---
title: "IBLI evaluation"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

## Packages

```{r lib, message=FALSE}
rm(list = ls(all=TRUE))
library(dplyr)
library(raster)
```

## Data preparation

### Administrative data

Load sub-location boundaries

```{r d12}
subloc <- agrodata::data_ibli("marsabit")
names(subloc)[6] <- "sublocation"
subloc
```

### NDVI and rainfall data.

Get Normalized Difference Vegetative Index (NDVI) data from NOAA-AVHRR, MODIS and rainfall from CHIRPS. The datasets are already spatially aggregated (mean) with respect to sublocation boundaries.

```{r d1}
n <- agrodata::data_ibli("marsabit_avhrr_ndvi") #NOAA
colnames(n)[1:5]
m <- agrodata::data_ibli("marsabit_modis_ndvi_agg") #MODIS
colnames(m)[1] <- toupper(colnames(m)[1])
colnames(m)[1:5]
p <- agrodata::data_ibli("marsabit_chirps_precipitation") #rainfall
colnames(p)[1:5]
```


Reshape the data from wide to long with date as a variable. 

```{r d2}
p <- reshape2::melt(p, variable.name = "date", value.name = "rainfall",id.vars = "SUBLOCATION")
p[1,]
n <- reshape2::melt(n, variable.name = "date", value.name = "noaa", id.vars = "SUBLOCATION")
n[1,]
m <- reshape2::melt(m, variable.name = "date", value.name = "modis", id.vars = "SUBLOCATION")
m[1,]

```

Format the date variable by removing unnecessary variables.

```{r d3}
#Chirps
p$date <- as.Date(p$date, format = "X%Y%m%d")
p[1,]
#NOAA
n$date <- as.Date(n$date, format = "X%Y%m%d")
n[1,]
#MODIS
m$date <- as.Date(m$date, format = "X%Y_%m_%d")
m[1,]

```
Compute daily average rainfall and NDVI per season in each year. First Create a function to compute means for Long Rain Long Dry (March -- September) and Short Rain Short Dry (October -- February) seasons.

```{r fxn1}
seasonMean <- function(year, df, seasons=1:2) {
	#df$date <- as.Date(df$date, format = "X%Y%m%d")
  res <- list()
  for (i in seasons) {
    season <- ifelse(i==1, "long", "short")
    if (season =="long") {
      sdate <- paste0(year, "-03-01")
      edate <- paste0(year, "-09-30")
      season <- "LRLD"
    } else if (season =="short") {
      sdate <- paste0(year-1, "-10-01")
      edate <- paste0(year, "-02-28")
      season <- "SRSD"
    } else {
      stop("Define season")
    }
    ydf <- df[df$date >= sdate & df$date <= edate, ]
    ym <- aggregate(ydf[,3], ydf[,1, drop=FALSE], mean, na.rm=T)
    ym$year <- year
    ym$season <- season
    res[[i]] <- ym    
  }  
  do.call(rbind, res)  
}

```

Now compute the averages using the function.

```{r d4}
years <- 1982:2019
#CHIRPS
temp <- lapply(years, seasonMean, p)
p <- do.call(rbind, temp)
names(p)[2] <- "rainfall"

#NOAA NDVI
temp <- lapply(years, seasonMean, n)
n <- do.call(rbind, temp)
names(n)[2] <- "noaa"

#MODIS NDVI
years <- 2001:2019
temp <- lapply(years, seasonMean, m)
m <- do.call(rbind, temp)
names(m)[2] <- "modis"
```


### Livestock data

Mortality and causes of mortality data

```{r d13}
mort <- agrodata::data_ibli("marsabit_mortality.rds")
colnames(mort)[colnames(mort)=="type"] <- "animal"
mort <- mort[mort$animal %in% c("Camel", "Cattle", "Shoat"), ]  # remote TLU for merging
mort <- mort[!is.na(mort$mortality_rate), ]                     # remove zero stock

cause <- agrodata::data_ibli("marsabit_losses")[,c("hhid", "sublocation", "animal", "cause","month", "year")]
cause$animal[cause$animal == "Goat/Sheep"] <- "Shoat"
cause$season <- "SRSD"
cause$season[cause$month >=3 & cause$month <=9] <-"LRLD"
head(cause, n=2)
cause <- na.omit(unique(cause))
table(cause$cause)
```

Merge causes table with mortality data.

We cannot really merge these. For now, to make progress, let's just drop that

```{r d3a}
nrow(mort)
nrow(cause)
mortx  <- merge(mort, cause, by=c("hhid","sublocation", "animal", "season", "year"), all.x=TRUE)
nrow(mortx)
```

Compute mortality

It might be of interest to do a separate model for Starvation/Drought


```{r d5}
#Consider: Disease, Starvation, Rain, predation
#  can also be stress induced
#mort2 <- mort2[!(mort2$cause %in% c("Just lost", "Old age", "Premature birth")),]
#mort2 <- mort2[mort2$type == "TLU", ]
#mort2 <- na.omit(mort2)

mort <- agrodata::data_ibli("marsabit_mortality.rds")
mort2 <- mort[mort$type == "TLU", ]  # remote TLU for merging

amort <- aggregate(mort2[,c("stock_beginning", "loss"), drop=FALSE], mort2[, c("season","year","sublocation"), drop=FALSE], mean, na.rm=TRUE)
amort$mortality_rate <- (amort$loss / amort$stock_beginning) * 100
```

#### Spatial Variation of mortality

Display the spatial variation in mortality.

```{r sv1}
temp <- aggregate(amort[, "mortality_rate", drop=FALSE], amort[, c("season", "sublocation"), drop=FALSE],mean, na.rm=T)
#LRLD
lr <- terra::merge(subloc, temp[temp$season=="LRLD", c("sublocation", "mortality_rate")], by= "sublocation")
#SRSD
sr <- terra::merge(subloc, temp[temp$season=="SRSD", c("sublocation", "mortality_rate")], by= "sublocation")
#ALL
all <- terra::merge(subloc, temp[, c("sublocation", "mortality_rate")], by= "sublocation")

cols <- heat.colors(length(lr$sublocation), rev=T)
sp.label <- function(x, label) {list("sp.text", coordinates(as(x, 'Spatial')), label, cex=0.5, col="black")}
NUMB.sp.label <- function(x) {sp.label(x, as.vector(x$sublocation))}
make.NUMB.sp.label <- function(x) {do.call("list", NUMB.sp.label(x))}
```

```{r spplot1}

raster::spplot(lr, "mortality_rate", col.regions=cols,
                    sp.layout = make.NUMB.sp.label(sr), 
                    cex = 0.5,
                    bg = "white", col="light grey", border="light grey",
                    main="LRLD average mortality")
```

```{r spplot2}
raster::spplot(sr, "mortality_rate", col.regions=cols,
                    sp.layout = make.NUMB.sp.label(sr), 
                    cex = 0.5,
                    bg = "white", col="light grey", border="light grey",
                    main="SRSD average mortality")
```

```{r spplot3}

raster::spplot(all, "mortality_rate", col.regions=cols,
                    sp.layout = make.NUMB.sp.label(all), 
                    cex = 0.5,
                    bg = "white", col="light grey", border="light grey",
                    main="Average mortality")
```

Mortality very high in LOIYANGALANI, LOGOLOGO and SAGANTE. In contrast it is Very low in KALACHA, TURBI, BUBISA (they should not take insurance).

Although there is variation by season, there seems to be support for grouping the large sublocations in the center and north east.

## Normal vs lognormal data treament 

Compute z-Score based data time series lengths. Z-score function.
,
```{r zscore}
zscore <- function(y, log=FALSE){
    # +.05 to avoid NA from log(x) where x <= 0
    if (log) y <- log(y+.05)
    (y - mean(y, na.rm=TRUE) ) / (sd(y, na.rm=TRUE))
}
```

Compute rainfall and NDVI Z-scores per season per sublocation.

```{r n2}
#Create data frame
df1 <- merge(p, n, by = c("SUBLOCATION","year", "season")) #NOAA and CHIRPSs
df1 <- merge(df1, m, by = c("SUBLOCATION","year", "season"), all.x=TRUE)#Add modis data
#Compute Z-scores per season per sublocation
dff <- ungroup(mutate(group_by(df1, SUBLOCATION, season), znoaa=zscore(noaa), zrain=zscore(rainfall), zmodis=zscore(modis),  zlnoaa=zscore(noaa, TRUE), zlrain=zscore(rainfall, TRUE), zlmodis=zscore(modis, TRUE)))

```

Check predictors distributions.

```{r nn1, message=FALSE}
library(MASS)
distfun <- function(x, main="") {
   x <- na.omit(x)
   #par(mfrow=c(1,2))
   for (i in 1:2) {
     if(i==2) x <- na.omit(log(x+0.05))
     hist(x, breaks=30, prob=TRUE, xlab="index", main = ifelse(i==1, main, paste0("log(",main,")")))
     fit <- fitdistr(x, "normal")
     curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col = "red", lwd=2, add = TRUE)
   }
}
```

Data distribution. 

```{r nn11, message=FALSE}
x11()
par(mfrow=c(2, 3), mar=c(4.5, 3.8, 1.8, 1)) #c(bottom, left, top, right)
#NOAA
distfun(na.omit(dff$noaa), main="NOAA")
#MODIS NDVI
distfun(dff$modis, main="MODIS")
#CHIRPS
distfun(dff$rainfall, main="Rainfall")
```

It seems the *log* transforms the rainfall data to a distribution close to normal. We will thus compute zScored log of transformed daily average rainfall and NDVI per season per sub-location.


Merge mortality data with NDVI and rainfall data.

```{r d6}
colnames(amort)[3] <- "SUBLOCATION"
dff <- merge(dff, amort, by=c("year", "season", "SUBLOCATION"))
```


## Mortality modelling and prediction

We will determine the effect rainfall data transformation on mortality predictions based on R$^2$ and Root Mean Square Error (RMSE). RMSE and R$^2$ functions are given as:

This is the relative RMSE (relative to a null model. that is more informative)
```{r fxn2}
rrmse <- function(observed, predicted){ 
  error <- observed - predicted
  rmse <- sqrt(mean(error^2, na.rm=TRUE))
  nullerror <- observed - mean(observed)
  nullrmse <- sqrt(mean(nullerror^2, na.rm=TRUE))
  (nullrmse - rmse) / nullrmse
}
```

I now see that this is actually the same as the above! 

```{r fxn22}
Rsq <- function(actual, predicted) { 
  d <- na.omit(data.frame(actual, predicted))
  1 - (sum((d$actual-d$predicted)^2)/ sum((d$actual - mean(d$actual))^2)) 
} 
```

### Regime switching linear regression and segmented regression

Here we test the performance of piecewise linear regression model and segmented regression in predicting livestock mortality. Piecewise will consider two  regimes; those when conditions are below zero and -0.5. Basically, we care more on the extreme cases because that is when farmers should receive compensation. We also evaluate farmers welfare benefit based on MQS with the assumption that 1TLU=1000$ and that each farmer has 1TLU before any shocks. 

```{r p1}
library(segmented)
#MQS function
library(agro)
mqs_test <- function(rho, df){
  ce_base <- ce_income(df$capital, rho)
  ce_ins  <- ce_income(df$income_ins, rho)
  ce_ins - ce_base
}

get_benefit <- function(df, trigger, rho=1.5) {
  #Define trigger
  data <- na.omit(df)
  #trigger <- quantile(data$predicted_mortality, 0.80, na.rm=T)
  #Compute payouts
  data$payouts <- pmax(0, data$predicted_mortality - trigger)/100 * 1000
  #Compute Actuarily fair premiumL Premium with no mark up
  premium <- mean(data$payouts, na.rm=T)
  # markup 25%
  premium <- 1.25 * premium
  #Compute income with insurance
  data$income_ins   <- (data$capital + data$payouts) - premium
  #Determine insurance benefit
  mqs_test(rho, data)
}
```

Regression

```{r p11}
regfun <- function(x, y, triggers, main="") {
  df <- na.omit(data.frame(x=x, y=y))
  df0 <- df[df$x<0,]
  df5 <- df[df$x < -0.5,]
  test <- data.frame(matrix(nrow=4, ncol=6+length(triggers)))
  mqsn <- paste0("mqs", triggers)
  nt <- length(triggers)
  colnames(test) <- c("group", "model", "r2", "RMSE", "intercept", "slope", mqsn)
  test$group <- main
  test$model <- c("lm", "lm0", "lm5", "seg")
  plot(y~x, pch=16, main=main, cex=0.9, cex.main=1.2, cex.lab=1.2)

  ml <- lm(y~x, data=df)
  abline(ml, col="magenta", lwd=3, lty=2)  
  ml2 <- lm(y~x, data=df0)
  abline(ml2, col="blue", lwd=3, lty=3)  
  ml5 <- lm(y~x, data=df5)
  abline(ml5, col="green", lwd=3, lty=4)  

  sm <- segmented(ml, seg.Z = ~x, psi=0)
  plot.segmented(sm, add=T, lwd=2, lty=1, col="red") 
  
  df$capital <- (100 - df$y)/100 * 1000
  e <- predict(ml)
  df$predicted_mortality <- e
  
  for (i in 1:nt) test[1,i+6] <- get_benefit(df, triggers[i])
  test[1,3:4] <- c(Rsq(df$y, e), rrmse(df$y, e))
  test[1,5:6] <- coefficients(ml)
  
  e <- predict(ml2)
  df$predicted_mortality <- NA
  df$predicted_mortality[df$x<0] <- e
  for (i in 1:nt) test[2,i+6] <- get_benefit(df, triggers[i])

  test[2,3:4] <- c(Rsq(df0$y, e), rrmse(df0$y, e))
  test[2,5:6] <- coefficients(ml2)
  
  e <- predict(ml5)
  df$predicted_mortality <- NA
  df$predicted_mortality[df$x < -0.5] <- e
  for (i in 1:nt) test[3,i+6] <- get_benefit(df, triggers[i])
  test[3,3:4] <- c(Rsq(df5$y, e), rrmse(df5$y, e))
  test[3,5:6] <- coefficients(ml5)
  
  e <- predict(sm)
  df$predicted_mortality <- e
  for (i in 1:nt) test[4,i+6] <- get_benefit(df, triggers[i])
  
  test[4,3:4] <- c(Rsq(df$y, e), rrmse(df$y, e))
  segslope <- slope(sm)$x[1]
  bx = sm$psi[2]
  by = predict(sm, data.frame(x=-0.92))
  intercpt = segslope * -bx + by
  test[4,5:6] <- c(intercpt, segslope)
  
  legend("topright", c("lm", "lm0", "lm5", "sm"), col=c("magenta", "blue", "green", "red"), lwd=c(3,3,3,2), lty=c(2,3,4,1), title="Model", bty="n")
  test
}

```


Possible pay-out triggers

```{r p2}
mort <- na.omit(dff$mortality_rate)
trigs <- round(quantile(mort, seq(0.6, 0.95, 0.05)), 1)
trigs

aggregate(dff[, "mortality_rate", drop=FALSE], dff[,"SUBLOCATION", drop=FALSE], function(i) quantile(i, c(0.75, 0.8), na.rm=T))
```

Regression models.

```{r p3}
par(mfrow=c(2, 3), mar=c(4.5, 3.6, 1.8, 1)) #c(bottom, left, top, right)
res <- list()
#NOAA NDVI and Log transformed NOAA NDVI.
res$noaa <- regfun(dff$znoaa, dff$mortality_rate, trigs, "NO")
res$lognoaa <- regfun(dff$zlnoaa, dff$mortality_rate, trigs, "lNO")
#CHIRPS rainfall and log transformed CHIRPS rainfall.
res$rain <- regfun(dff$zrain, dff$mortality_rate, trigs, "RN")
res$lograin <- regfun(dff$zlrain, dff$mortality_rate, trigs, "lRN")
#MODIS NDVI and log transformed MODIS NDVI.
res$modis <- regfun(dff$zmodis, dff$mortality_rate, trigs, "MD")
res$logmodis <- regfun(dff$zlmodis, dff$mortality_rate, trigs, "lMD")
```


```{r p6}
x <- do.call(rbind, res)
rownames(x) <- NULL
x[,3:ncol(x)] <- lapply(x[,3:ncol(x)], function(i) round(i,3))
x
```

#### How good would a perfect model be?

```{r}
dff$capital <- (100 - dff$mortality_rate)/100 * 1000
dff$predicted_mortality <- dff$mortality_rate
pf <- sapply(trigs, function(i) get_benefit(dff, i)) 
pf
```

Compute a relative IB score `RIB`. 

```{r}
xx <- x
for (i in 7:ncol(x)) {
  xx[,i] <- x[,i] / pf[i-6]
}

apply(xx[,7:ncol(xx)], 2, max)
```


# save data 

Save data for use in next steps.

```{r p7}
#x <- x[x$model != "loess", ]
saveRDS(x,  "regression_models.rds")
saveRDS(pf, "perfect.rds")
saveRDS(xx, "reg_RIB.rds")
```


