---
title: "Experiments for ideas of 25 June 2020"
author: "Benson Kenduiywo"
date: "26/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

Please install *raster* ` `, terra `remotes::install_github("rspatial/terra")` and *agrodata* `remotes::install_github("reagro/agrodata")` in that order. Load necessary libraries.


```{r lib, message=FALSE}
rm(list = ls(all=TRUE))
library(agrodata)
library(reshape2)
library(dplyr)
```

## Data preparation

### Administrative data

Load sub-location boundaries

```{r d12}
subloc <- agrodata::data_ibli("marsabit")
names(subloc)[6] <- "sublocation"
subloc
```

### NDVI and rainfall data.

Get Normalized Difference Vegetative Index (NDVI) data from NOAA-AVHRR, MODIS and rainfall from CHIRPS. The datasets are already spatially aggregated (mean) with respect to sublocation boundaries.

```{r d1}
n <- agrodata::data_ibli("marsabit_avhrr_ndvi") #NOAA
colnames(n)[1:5]
m <- agrodata::data_ibli("marsabit_modis_ndvi_agg") #MODIS
colnames(m)[1] <- toupper(colnames(m)[1])
colnames(m)[1:5]
p <- agrodata::data_ibli("marsabit_chirps_precipitation") #rainfall
colnames(p)[1:5]
```


Reshape the data from wide to long with date as a variable. 

```{r d2}
p <- melt(p, variable.name = "date", value.name = "rainfall",id.vars = "SUBLOCATION")
p[1,]
n <- melt(n, variable.name = "date", value.name = "noaa", id.vars = "SUBLOCATION")
n[1,]
m <- melt(m, variable.name = "date", value.name = "modis", id.vars = "SUBLOCATION")
m[1,]

```

Format the date variable by removing unnecessary variables.

```{r d3}
#Chirps
date <- gsub("X", "", p$date)
p$date <- as.Date(as.character(date), format = "%Y%m%d")
p[1,]
#NOAA
date <- gsub("X", "", n$date)
n$date <- as.Date(as.character(date), format = "%Y%m%d")
n[1,]
#MODIS
date <- gsub("X", "", m$date)
m$date <- as.Date(as.character(date), format = "%Y_%m_%d")
m[1,]

```
Compute daily average rainfall and NDVI per season in each year. First Create a function to compute means for Long Rain Long Dry (March -- September) and Short Rain Short Dry (October -- February) seasons.

```{r fxn1}
seasonMean <- function(year, df, season){
	df$date <- as.Date(df$date, format = "X%Y%m%d")
  if(season =="long"){
    sdate <- paste0(year, "-03-01")
    edate <- paste0(year, "-09-30")
    season <- "LRLD"
  }else if (season =="short"){
    sdate <- paste0(year-1, "-10-01")
    edate <- paste0(year, "-02-28")
    season <- "SRSD"
  }else{
    print("Define season")
  }
  ydf <- df[df$date >= sdate & df$date <= edate, ]
  ym <- aggregate(ydf[,3], ydf[,1, drop=FALSE], mean, na.rm=T)
  ym$year <- year
  ym$season <- season
  return(ym)
}

```

Now compute the averages using the function.

```{r d4}
years <- seq(1982, 2019, 1)
#CHIRPS
temp1 <- lapply(years, seasonMean, p, "long")
temp1 <- do.call(rbind, temp1)
temp2 <- lapply(years, seasonMean, p, "short")
temp2 <- do.call(rbind, temp2)
p <- rbind(temp1, temp2)
names(p)[2] <- "rainfall"
#NOAA NDVI
temp1 <- lapply(years, seasonMean, n, "long")
temp1 <- do.call(rbind, temp1)
temp2 <- lapply(years, seasonMean, n, "short")
temp2 <- do.call(rbind, temp2)
n <- rbind(temp1, temp2)
names(n)[2] <- "noaa"
#MODIS NDVI
years <- seq(2001, 2019, 1)
temp1 <- lapply(years, seasonMean, m, "long")
temp1 <- do.call(rbind, temp1)
temp2 <- lapply(years, seasonMean, m, "short")
temp2 <- do.call(rbind, temp2)
m <- rbind(temp1, temp2)
names(m)[2] <- "modis"

```

### Livestock data

Mortality and causes of mortality data

```{r d13}
mort <- agrodata::data_ibli("marsabit_mortality.rds")
cause <- agrodata::data_ibli("marsabit_losses")[,c("hhid", "sublocation", "cause","month", "year")]
cause$season <- "SRSD"
cause$season[cause$month >=3 & cause$month <=9] <-"LRLD"
head(cause, n=2)
cause <- na.omit(cause)

```

Merge causes table with mortality data.

```{r d3a}
mort2  <- merge(mort, cause, by=c("hhid","sublocation", "season", "year"))

```
Compute mortality

```{r d5}
#Consider: Disease, Starvation, Rain, predation
mort2 <- mort2[!(mort2$cause %in% c("Just lost", "Old age","Premature birth")),]
mort2 <- mort2[mort2$type == "TLU", ]
mort2 <- na.omit(mort2)
amort <- aggregate(mort2[,c("stock_beginning", "loss"), drop=FALSE], mort2[, c("season","year","sublocation"), drop=FALSE], mean, na.rm=TRUE)
amort$mortality_rate <- (amort$loss/amort$stock_beginning) * 100

```

## Normal vs lognormal data treament 

compute z-Score based data time series lengths. Z-score function.

```{r zscore}
zscore <- function(y){ 
  (y - mean(y, na.rm=TRUE) ) / (sd(y, na.rm=TRUE))
  }

```

Compute Z-scores.

```{r n2}
df1 <- merge(p, n, by = c("SUBLOCATION","year", "season")) #NOAA and CHIRPSs
df1 <- ungroup(mutate(group_by(df1, SUBLOCATION, season), znoaa=zscore(noaa), zlograin=zscore(log(rainfall)), zrain=zscore(rainfall)))
#MODIS
df2 <- ungroup(mutate(group_by(m, SUBLOCATION, season), zmodis=zscore(modis)))

```

Check predictors distributions.

```{r n1, message=FALSE}
library(MASS)
#NOAA
x <- df1$noaa
x11()
hist(x, breaks=30, prob=T, xlab="NOAA NDVI", main = "NOAA NDVI")
fit <- fitdistr(x, "normal")
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col = 2, add = TRUE)
#MODIS NDVI
x <- df2$modis
x11()
hist(x, breaks=30, prob=T, xlab="MODIS NDVI", main = "MODIS NDVI")
fit <- fitdistr(x, "normal")
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col = 2, add = TRUE)
#CHIPRS
x <- df1$rainfall
x11()
hist(x, breaks=40, prob=T, xlab="CHIPRS Rainfall", main = "Rainfall")
fit <- fitdistr(x, "normal")
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col = 2, add = TRUE)

x11()
x <- log(df1$rainfall)
hist(x, breaks=40, prob=T, xlab="CHIPRS Rainfall", main = "Log transformed Rainfall")
fit <- fitdistr(x, "normal")
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col = 2, add = TRUE)

```

It seems the *log* transforms the rainfall data to a distribution close to normal. We will thus compute zScored log of transformed daily average rainfall and NDVI per season per sub-location.

Merge mortality with NDVI and rainfall data.

```{r d6}
colnames(amort)[3] <- "SUBLOCATION"
df1 <- merge(df1, amort, by=c("year", "season", "SUBLOCATION"))
df2 <- merge(df2, amort, by=c("year", "season", "SUBLOCATION"))

```

### Effect rainfall data transformation on mortality predictions

We will determine the effect rainfall data transformation on mortality predictions based on R$^2$, Mean Absolute Prediction Error (MAPE) and RMSE. For instance, a MAPE of (1) < 10 indicates highly accurate prediction, while (2) 10--0 good prediction, 3) 20 -- reasonable prediction, and (4) > 50 inaccurate.

MAPE, RMSE and R$^2$ functions are given as:

```{r fxn2}
rmse <- function(error){ return(sqrt(mean(error^2)))}
MAPE <- function (y_pred, y_true){ mean(abs((y_true - y_pred)/y_true)) * 100 }
R_square <- function(actual, predicted) { 1 - (sum((actual-predicted)^2)/sum((actual-mean(actual))^2))} 
```

Mortality modelling and prediction

```{r d7}
#Zscores from untransformed rainfall
m_rz <- loess(mortality_rate ~ zrain , data=df1, control=loess.control(surface="direct"))
e <- predict(m_rz, df1)
R_square(df2$mortality_rate, e)
rmse(df2$mortality_rate - e)
#log Transformed rainfall
m_rz <- loess(mortality_rate ~ zlograin , data=df1, control=loess.control(surface="direct"))
e <- predict(m_rz, df1)
R_square(df2$mortality_rate, e)
rmse(df2$mortality_rate - e)

```

## Short verss long time series computed Z-scores

Here we test the effect of shortening the time period of computing Z-scores from rainfall, NOAA AVHRR and MODIS data on mortality predictions. Therefore, we shorten the time series by leaving out a year up to the year 2008. This because we have observed mortality data from 2008--2015.

```{r t1}

#Impact of rainfal predictions based on shortening period of zScore calculations by 4 years starting from 1982
df <- merge(p, n, by = c("SUBLOCATION","year", "season"))
R2c <- RMSEc <- mapec <- R2b <- RMSEb <- mapeb <- R2 <- RMSE <- mape <- NULL
year <- seq(1982,2008,1)
i <- j <- 0
for(y in year){
  i=i+1 
  df4 <- df[df$year >= y,]
  df4 <- ungroup(mutate(group_by(df4, SUBLOCATION, season), znoaa=zscore(noaa), zrain=zscore(rainfall)))
  df4 <- merge(df4, amort, by=c("year", "season", "SUBLOCATION"))
  #Rainfall
  m_rz <- loess(mortality_rate ~ zrain , data=df4, control=loess.control(surface="direct"))
  e <- predict(m_rz, df4)
  R2[i] <- R_square(df4$mortality_rate, e)
  RMSE[i] <- rmse(df4$mortality_rate - e)
  mape[i] <- MAPE(df4$mortality_rate, e)
  #NOAA
  m_nz <- loess(mortality_rate ~ znoaa, data=df4, control=loess.control(surface="direct"))
  e <- predict(m_nz, df4)
  R2b[i] <- R_square(df4$mortality_rate, e)
  RMSEb[i] <- rmse(df4$mortality_rate - e)
  mapeb[i] <- MAPE(df4$mortality_rate, e)
  #MODIS NDVI
  if(y >=2001){
    j=j+1 
    df5 <- m[m$year >= y,]
    df5 <- ungroup(mutate(group_by(df5, SUBLOCATION, season), zmodis=zscore(modis)))
    df5 <- merge(df5, amort, by=c("year", "season", "SUBLOCATION"))
    m_mz <- loess(mortality_rate ~ zmodis, data=df5, control=loess.control(surface="direct"))
    e <- predict(m_mz, df5)
    R2c[j] <- R_square(df5$mortality_rate, e)
    RMSEc[j] <- rmse(df5$mortality_rate - e)
    mapec[j] <- MAPE(df5$mortality_rate, e)
  }
  
}

x11()
plot(R2~year, pch=1, xaxt = "n", ylab=expression(R^2), ylim=c(0.2,0.35), main="Impact of Z-score computation period in mortality predictions")
points(R2b~year, pch=3)
points(R2c~year[year>=2001], pch=16)
axis(side=1, at=year)
legend("topleft", pch = c(1, 3, 16),  legend=c('Rainfall','NOAA NDVI', "MODIS NDVI"), title="Model type")

```

## Regime switching linear regression vs lowess

Here we test a piecewise linear regression model against the lowess model. Piecewise will consider two climate regimes; those when conditions are below zero and above it.

```{r p1}
x=df1$znoaa
y=df1$mortality_rate
x11()
plot(x,y, pch=16)
pcs1 <- lm(y ~ x*(x <= -1) + x*(x >= -1))
summary(pcs1)
coef <- summary(pcs1)
intercept_a <- coef$coefficients[1,1] + coef$coefficients[2,1]
slope_a <- (coef$coefficients[2,1] + coef$coefficients[3,1]) 
curve(slope_a*x + intercept_a, add=T, from=-3, to=0)
###2 not defined because of singularities CHALLANGEE
```





