---
title: "zones"
date: "7/2/2020"
output: html_document
---
# Effect No zone vs group of sub-locations on MQS

## Clustering zones

Use kmeans to define/create 4 insurance zones based on spatial-temporal sublocation indices from NOAA NDVI, MODIS NDVI and rainfall indices per season per year.

```{r z1}
temp1 <- reshape2::dcast(df, SUBLOCATION ~ year+season, value.var="rainfall")
temp2 <- reshape2::dcast(df, SUBLOCATION ~ year+season, value.var="noaa")
temp3 <- reshape2::dcast(m, SUBLOCATION ~ year+season, value.var="modis")
df_w <- merge(temp1,temp2, by=c("SUBLOCATION"))
df_w <- merge(df_w, temp3, by=c("SUBLOCATION"))
subs <- sort(unique(amort$SUBLOCATION))
aoi <- subloc[subloc$sublocation %in% subs,]
df_w <- df_w[df_w$SUBLOCATION %in% subs,]
set.seed(99)
#rf <-  randomForest(df_w[,-1],ntree=200,importance = TRUE)
#Create 10 clusters using 500 iterations using 5 random sets from "Lloyd" method 
r_kmn <- kmeans(df_w[,-1], centers = 3, iter.max = 500, nstart = 1, algorithm="Lloyd")
# kmeans returns an object of class "kmeans"
aoi$Zones <- r_kmn$cluster

```

Display the old and new zones

```{r z2, message=FALSE}
plot(aoi, "Zones", main= "NDVI+Rain")
```

Merge to current boundaries

```{r zz2, message=FALSE}
z_10 <- aggregate(aoi, by="Zones")
plot(z_10, "Zones", border = "red", lwd = 2)
legend("bottomleft", c("Original", "Clustered"), col=c("black","red"), lty = 1, lwd=2, title="Sublocation clusters")
```

### MQS test

Prepare data for modelling and MQS

```{r z3}
#No zones defined
data1 <- ungroup(mutate(group_by(df, SUBLOCATION, season), znoaa=zscore(noaa), zrain=zscore(rainfall)))
temp <- ungroup(mutate(group_by(m, SUBLOCATION, season), zmodis=zscore(modis)))
data1 <- merge(data1, amort, by=c("SUBLOCATION", "season", "year"))
data1 <- merge(data1, temp, by=c("SUBLOCATION", "season", "year"))
#Add Zones
names(aoi)[6] <- "SUBLOCATION"
data1 <- merge(data1, as.data.frame(aoi[,c("SUBLOCATION", "Zones")]), by="SUBLOCATION")
#Zones defined
data2 <- merge(df, as.data.frame(aoi[,c("SUBLOCATION", "Zones")]), by="SUBLOCATION")
data2 <- ungroup(mutate(group_by(data2, Zones, season), znoaa=zscore(noaa), zrain=zscore(rainfall)))
temp <- merge(m, as.data.frame(aoi[,c("SUBLOCATION", "Zones")]), by="SUBLOCATION")
temp <- ungroup(mutate(group_by(temp, Zones, season), zmodis=zscore(modis)))

data2 <- merge(data2, amort, by=c("SUBLOCATION", "season", "year"))
data2 <- merge(data2, temp, by=c("SUBLOCATION", "season", "year", "Zones"))

```

Predict livestock mortality using NDVI (NOAA and MODIS) and rainfall.

```{r z4}
#RAINFALL ONLY
m_l <- loess(mortality_rate ~ zrain, data=data1, control=loess.control(surface="direct"))
p <- predict(m_l, data1)
r2 <- R_square(data1$mortality_rate, p)
cat("Rainfall model R2 values is ", r2, "\n")
#RAINFALL and NOAA NDVI
m_l <- loess(mortality_rate ~ zrain + znoaa, data=data1, control=loess.control(surface="direct"))
p <- predict(m_l, data1)
r2 <- R_square(data1$mortality_rate, p)
cat("Rainfall & NOAA NDVI model R2 values is ", r2, "\n")
#RAINFALL, NOAA NDVI and MODIS NDVI
m_l <- loess(mortality_rate ~ zrain + znoaa + zmodis, data=data1, control=loess.control(surface="direct"))
p <- predict(m_l, data1)
r2 <- R_square(data1$mortality_rate, p)
cat("Rainfall, NOAA NDVI & MODIS NDVI model R2 values is ", r2, "\n")
data1$predicted_mortality <- p

#Predict mortality with defined zones
data2 <- ungroup(mutate(group_by(data2, year, season, Zones),
                       mortality_rate=mean(mortality_rate)))
m_l <- loess(mortality_rate ~ zrain + znoaa + zmodis, data=data2, control=loess.control(surface="direct"))
p <- predict(m_l, data2)
r2 <- R_square(data2$mortality_rate, p)
cat("Rainfall, NOAA NDVI & MODIS NDVI model R2 values is ", r2, "\n")
data2$predicted_mortality <- p

```

Compute payouts — in units of monetary value of TLU — as the percentage difference between predicted mortality in TLU and a trigger (the 80th percentile). Those below the trigger get no payout. We assume that 1 TLU has a monetary equivalence of 1000$.

```{r z5}
trig <- quantile(data1$predicted_mortality, 0.80, na.rm=T)
data1$payouts <- pmax(0, data1$predicted_mortality - trig)/100 * 1000
#Case with defined zones
trig <- quantile(data2$predicted_mortality, 0.80, na.rm=T)
data2$payouts <- pmax(0, data2$predicted_mortality - trig)/100 * 1000

```

Assume an actuarially fair premium contract and farmers capital the percentage of herd remaining after loss (observed mortality) with 1TLU=1000$. 

```{r z6}
premium <- mean(data1$payouts)
cat("The premium is $", premium,"\n",sep = "")
data1$capital <- ((100 - data1$mortality_rate)/100) * 1000
data1$income_noins <- data1$capital
data1$income_ins   <- (data1$capital + data1$payouts) - premium
#Case with defined zones
premium <- mean(data2$payouts)
data2$capital <- ((100 - data2$mortality_rate)/100) * 1000
data2$income_noins <- data2$capital
data2$income_ins   <- (data2$capital + data2$payouts) - premium

```

Let us now evaluate insurance welfare with insurance zones and without

```{r mqs1}
library(agro)
test <- function(rho, df1){
  ce_base <- ce_income(df1$income_noins, rho)
  ce_ins  <- ce_income(df1$income_ins, rho)
  mqs    <- ce_ins - ce_base
  return(mqs)
}
zoned <- unzoned <- NULL

rhos <- seq(0, 10, 1)
for(z in 1:length(rhos)){
  #CASE 1: NO ZONES
  unzoned[z] <- test(rhos[z], data1)
  #CASE 2: DEFINED INSURANCE ZONES
  zoned[z]   <- test(rhos[z], data2)
}

#plot(zoned~rhos, ylab="Insurance benefit", xlab="Risk aversion", pch=16, col="blue")
#points(unzoned~rhos,pch=16,col="red")
#legend("bottomright",   legend=c("Unzoned",  "Zoned"), fill=c("red", "blue"), title=expression(paste("Region type")), bty = "n")
#zoned
#unzoned

```


