---
title: "IBLI contract design and evaluation based on chirps and NDVI"
author: "Benson Kenduiywo"
date: "22/11/2019"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## IBLI Programme Design and Evaluation using CHIRPS rainfall data

Load libraries
 
```{r lib}
rm(list = ls(all=TRUE))
library(agrins)
library(reshape2)
library(dplyr)

```

Load precipitation and NDVI data.

```{r d1}
n <- data_ibli("marsabit_avhrr_ndvi") #NOAA
p <- data_ibli("marsabit_chirps_precipitation")
names(p)[1:10]

```

Reshape the data from wide table to long with year as variable.

```{r d2}
p2 <- melt(p, variable.name = "date", value.name = "rainfall", id.vars = "SUBLOCATION")
p2[1,]

n2 <- melt(n, variable.name = "date", value.name = "ndvi", id.vars = "SUBLOCATION")
n2[1,]
```

Format the date variable by removing unnecessary variables.

```{r d3}
#Chirps
date <- gsub("X", "", p2$date)
p2$date <- as.Date(as.character(date), format = "%Y%m%d")
p2[1,]
#NOAA
date <- gsub("X", "", n2$date)
n2$date <- as.Date(as.character(date), format = "%Y%m%d")
n2[1,]

```

Let us make some plots.

```{r plots}
plot(rainfall~date, data=p2, type="l", ylab="rainfall (mm)")
m <- months(p2$date)
i <- m == "September"
lines(p2$date[i], p2$rainfall[i], col="red", cex=1.2)
j <- m == "October"
lines(p2$date[j], p2$rainfall[j], col="blue", cex=1.2)
legend("topleft", lty = 1, col = c("red", "blue"), legend=c('September','October'), title="Month")

#Normalize data function (0-1)
nor <- function(x){
  (x - min(x))/ (max(x)-min(x))
}

#1982-2019 average annual rainfall
p2$year <- as.numeric(format(p2$date,'%Y')) 
y <- aggregate(rainfall~year, p2, mean, na.rm=T)
ref <- y
x11()
plot(rainfall~year, data=y[-1,], type="l", ylab="rainfall (mm)", main="1982-2019 average annual rainfall", axes=F, cex.lab=1.2)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, 5, by=1), cex.axis=1.1)
box()

#1982-2019 average annual NDVI
n2$year <- as.numeric(format(n2$date,'%Y')) 
y <- aggregate(ndvi~year, n2, mean, na.rm=T)
ref$ndvi <- y$ndvi
x11()
plot(ndvi~year, data=y[-1,], type="l", ylab="NDVI", main="1982-2019 average annual NDVI", axes=F, cex.lab=1.2)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, 0.5, by=0.1), cex.axis=1.1)
box()

```

Create a function to compute means for Long Rain Long Dry (March -- September) and Short Rain Short Dry (October -- February) seasons.

```{r d4}
seasonMean <- function(year, df, season){
  if(season =="long"){
    sdate <- paste0(year, "-03-01")
    edate <- paste0(year, "-09-30")
    season <- "LRLD"
  }else if (season =="short"){
    sdate <- paste0(year-1, "-10-01")
    edate <- paste0(year, "-02-28")
    season <- "SRSD"
  }else{
    print("Define season")
  }
  ydf <- df[df$date >= sdate & df$date <= edate, ]
  ym <- aggregate(ydf[,3], ydf[,1, drop=FALSE], mean, na.rm=T)
  ym$year <- year
  ym$season <- season
  return(ym)
}

```

Compute daily average rainfall per season in each year.

```{r d5}
years <- seq(1982, 2019, 1)
LR_rain <- lapply(years, seasonMean, p2, "long")
LR_rain <- do.call(rbind, LR_rain)
names(LR_rain)[2] <- "rainfall"
SR_rain <- lapply(years, seasonMean, p2, "short")
SR_rain <- do.call(rbind, SR_rain)
names(SR_rain)[2] <- "rainfall"
rain <- rbind(LR_rain, SR_rain)

#1982-2019 average annual rainfall per season
y <- aggregate(rainfall~year+season, rain, mean, na.rm=T)
x11()
plot(rainfall~year, data=y, type="n", ylab="rainfall (mm)", main="1982-2019 average annual rainfall per season",  axes=FALSE, cex.lab=1.2)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, 5, by=1), cex.axis=1.1)
box()
i <- y[y$season=="LRLD", ]
lines(rainfall~year, i, col="red", cex=1.2)
i <- y[y$season=="SRSD", ]
lines(rainfall~year, i, col="blue", cex=1.2)
legend("topleft", lty = 1, col = c("red", "blue"), legend=c('LRLD','SRSD'), title="Season")

#1982-2019 average annual NDVI per season
LR_ndvi <- lapply(years, seasonMean, n2, "long")
LR_ndvi <- do.call(rbind, LR_ndvi)
names(LR_ndvi)[2] <- "ndvi"
SR_ndvi <- lapply(years, seasonMean, n2, "short")
SR_ndvi <- do.call(rbind, SR_ndvi)
names(SR_ndvi)[2] <- "ndvi"
ndvi <- rbind(LR_ndvi, SR_ndvi)
y <- aggregate(ndvi~year+season, ndvi, mean, na.rm=T)
x11()
plot(ndvi~year, data=y, type="n", ylab="NDVI", main="1982-2019 average annual NDVI per season",  axes=FALSE, cex.lab=1.2)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, 1, by=0.1), cex.axis=1.1)
box()
i <- y[y$season=="LRLD", ]
lines(ndvi~year, i, col="red", cex=1.2)
i <- y[y$season=="SRSD", ]
lines(ndvi~year, i, col="blue", cex=1.2)
legend("topleft", lty = 1, col = c("red", "blue"), legend=c('LRLD','SRSD'), title="Season")

```

Compute zScored average daily rainfall per season per Sub-location. First get the zScore function.

```{r d6}
zscore <- function(y){
  (y - mean(y, na.rm=TRUE) ) / (sd(y, na.rm=TRUE))
}

```

The compute the zScored daily average rainfall per season per sublocation.

```{r d7}
rain <- ungroup(mutate(group_by(rain,SUBLOCATION), 
                       zrain=zscore(rainfall)))

ndvi <- ungroup(mutate(group_by(ndvi,SUBLOCATION), 
                       zndvi=zscore(ndvi)))
```

Now we need household mortality data in order to build a model. Get mortality data.

```{r d8}
mort <- data_ibli("marsabit_mortality.rds")

```

Load data with causes of mortality.

```{r d9}
cause <- data_ibli("marsabit_losses")[,c("hhid", "sublocation", "cause","month", "year")]
colnames(cause)[4] <- "season"
s <- cause$season
cause$season <- "SRSD"
cause$season[s >=3 & s <=9] <-"LRLD"
cause <- na.omit(cause)
head(cause, n=2)

```

Merge causes table with mortality data.

```{r d10}
mort2  <- merge(mort, cause, by=c("hhid","sublocation", "season", "year"))

```

Make some plots of cause of deaths.

```{r d11}
temp <- mort2[mort2$type=="TLU", ]
c <- tapply(temp$loss, temp$cause, mean, na.rm=TRUE)
c <- aggregate(loss~cause+year, temp, mean, na.rm=T)
c <- reshape(c, direction="wide", idvar="cause", timevar="year")
colnames(c) <- gsub("loss.", "", colnames(c))
c$cause <- gsub("/.*","", c$cause)
c$cause <- gsub(" .*","", c$cause)
c$cause[c$cause=="Just"] = "Lost"
c <- as.matrix(c)
rownames(c) <- c[,1]
cola <- c("red", "green", "blue", "grey40", "yellow", "black", "orange", "purple", "brown", "skyblue")
par(mfrow=c(1, 1), mar=c(5, 5, 4, 5))
barplot(c[,-1], legend =  rownames(c), las=2, args.leg=list(x = "topright", bty = "n", inset=c(-0.2, 0),cex=1), col=cola, ylab="Livestock loss per TLU")

```

## Spatial Variation
 
Let us explore if there is any spatial variation of average rainfall since 1980.

Load Marsabit shapefile

```{r sp1}
subloc <- data_ibli("marsabit")
```

Check names consistency before merging.

```{r sp2}
a = sort(unique(subloc$SUB_LOCATION))
b = sort(unique(rain$SUBLOCATION))
a[!a %in% b]

```

Now format sublocation names in shapefile to be consistent with those in rain.

```{r sp3}
subloc$SUB_LOCATION <- gsub("\\s*\\([^\\)]+\\)","",subloc$SUB_LOCATION)
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MARSABIT TOWNSHIP"] <- "TOWNSHIP"
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MARSABIT TOWNSHIP"] <- "TOWNSHIP"
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MT. KULAL" ] <- "MT KULAL" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="LONYORIPICHAU" ] <- "LONYORI PICHAU" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="EL-HADI" ] <- "EL-HARDI" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="KURUNGU" ] <- "KARUNGU"

```

Now merge rainfall & NDVI data with sublocation boundary information and plot LRLD and SRSD spatial variation maps.

```{r sp4}
#rainfall
r <- aggregate(rain[, c("zrain","rainfall"), drop=FALSE], rain[,c("SUBLOCATION", "season"), drop=FALSE],mean, na.rm=T)
#ndvi
n3 <- aggregate(ndvi~SUBLOCATION+season, data=ndvi,mean, na.rm=T)
temp <- merge(n3, r, by=c("SUBLOCATION","season"))
#LRLD
lr <- merge(subloc, temp[temp$season=="LRLD",], by.x="SUB_LOCATION", by.y="SUBLOCATION")
#SRSD
sr <- merge(subloc, temp[temp$season=="SRSD",], by.x="SUB_LOCATION", by.y="SUBLOCATION")

cols <- terrain.colors(length(lr$SUB_LOCATION),rev = T)
#Rainfal spatial variation
x11()
spplot(lr, "rainfall", col.regions=cols, main="LRLD Average rainfal per Sub-location: 1982-2019")
x11()
spplot(sr, "rainfall", col.regions=cols, main="SRSD Average rainfall per Sub-location: 1982-2019")

#NDVI spatial variation
x11()
spplot(lr, "ndvi", col.regions=cols, main="LRLD Average NDVI per Sub-location: 1982-2019")
x11()
spplot(sr, "ndvi", col.regions=cols, main="SRSD Average NDVI per Sub-location: 1982-2019")

```

## Modelling Mortality

### Introduction

To design and evaluate an index based insurance contract, we need to understand the relationship between the index (rainfall and NDVI) and losses incurred. Here we develop such a relationship using the z-scored rainfall and NDVI to predict livestock mortality, using survey data from Marsabit.

To establish the relationship we need to estimate mortality for each location and season. We will use household seasonal location average mortality rate $M_{ls})$. However, since sample households varied among the different surveys we will rely on seasonal $s$ average herd mortality per household for each location, $\overline{M}_{ls}$, to compute $M_{ls}$ using

$$
  M_{ls} = \frac{\sum_{m\in s} \overline{M}_{ls} }{\max_{m\in s} \overline{H}_{ls}},
$$ 

where $\overline{H}_{ls}$ is the seasonal *s* (either LRLD or SRSD seasons) beginning herd size averaged per household for each location.

```{r mortality}
#Consider: Disease, Starvation, Rain, predation
#mort2 <- mort2[(mort2$cause %in% c("Disease", "Predation","Rain","Starvation/Drought")),]
mort2 <- mort2[!(mort2$cause %in% c("Just lost", "Old age","Premature birth")),]
mort2 <- mort2[mort2$type == "TLU", ]
amort <- aggregate(mort2[,c("stock_beginning", "loss"), drop=FALSE], mort2[, c("season","year","sublocation"), drop=FALSE], mean, na.rm=TRUE)
amort$mortality_rate <- amort$loss/amort$stock_beginning

```

Merge mortality with NDVI and rainfall data.

```{r m2}
colnames(ndvi)[1] <- "sublocation"
d <- merge(ndvi, amort, by=c("year", "season", "sublocation"))
colnames(rain)[1] <- "sublocation"
d <- merge(rain, d, by=c("year", "season", "sublocation"))

```

Explore the relation between mortality and NDVI.

```{r m3}
cols <- c("red", "blue")
seas <- c("SRSD", "LRLD")
pcols <- cols[match(d$season, seas)]
x11()
par(mfrow=c(2, 1), mar=c(4, 4, 1, 10)) #c(bottom, left, top, right)
plot(mortality_rate~zrain, data=d, xlab="Z-rainfall (mm)", ylab="mortality", col=pcols, pch=20)
legend("topright", seas, col=cols, pch=20)
#NDVI
#x11()
plot(mortality_rate~zndvi, data=d, xlab="ZNDVI", ylab="mortality", col=pcols, pch=20)
legend("topright", seas, col=cols, pch=20)

```

The relationship is a bit noisy (as to be expected). There is a lot of variability at *zNDVI < -1* and *zrain < -1*. That could be a problem; as predictions of mortality at low zNDVI would be quite uncertain. This is why we will do a simulation later.

### Mortality model

Build a local regression model. Note:  change the surface parameter in loess to "direct" to extrapolate values.

```{r m4, message=FALSE}
dd <- na.omit(d[, c("zndvi", "zrain", "mortality_rate")])
dd$mortality_rate <- dd$mortality_rate * 100
library(msir)
m <- loess.sd(dd$zndvi, dd$mortality_rate, control=loess.control(surface="direct"))

x11()
par(mfrow=c(1, 2), mar=c(15, 4, 2, 0.5)) #c(bottom, left, top, right)
#par(mfrow=c(1, 2))
plot(dd$zndvi, dd$mortality_rate, cex=0.5, col="blue", xlab="zNDVI", ylab="Mortality rate (%)", main="NDVI Mortality Model", cex.lab=1.2, cex.axis=1.15)
lines(m$x, m$y, col="red")
lines(m$x, m$upper, lty=2)
lines(m$x, m$lower, lty=2)

#rainfall
m <- loess.sd(dd$zrain, dd$mortality_rate, control=loess.control(surface="direct"))
#x11()
plot(dd$zrain, dd$mortality_rate, cex=0.5, col="blue", xlab="z-rainfall", ylab="Mortality rate (%)", main="Rainfall Mortality Model", cex.lab=1.2, cex.axis=1.2)
lines(m$x, m$y, col="red")
lines(m$x, m$upper, lty=2)
lines(m$x, m$lower, lty=2)

```

Compare predicted with observed mortality rates.

```{r m5}
#Rainfall
m_r <- loess(mortality_rate ~ zrain , data=dd, control=loess.control(surface="direct"))
p <- predict(m_r, dd)
par(mfrow=c(1, 2), mar=c(15, 4, 1.5, 0.5))
plot(dd$mortality_rate, p, xlab="Observed mortality (%)", ylab="Predicted mortality (%)", main="Rainfall model", pch=16, cex.axis=1.2, cex.lab=1.2, cex=0.7)
abline(0,1, lwd=2, col="red")
cat("Rainfall model Residual Standard Error: ", m_r$s, "\n")
#NDVI
m_n <- loess(mortality_rate ~ zndvi, data=dd, control=loess.control(surface="direct"))
p <- predict(m_n, dd)
plot(dd$mortality_rate, p, xlab="Observed mortality (%)", ylab="Predicted mortality (%)", main="NDVI model", pch=16, cex.axis=1.2, cex.lab=1.2, cex=0.7)
abline(0,1, lwd=2, col="red")
cat("NDVI model Residual Standard Error: ", m_n$s, "\n")

```

We can observe that NDVI & Rainfall model has the least residual Standard Error while NDVI model has the highest standard error indicating low prediction accuracy. We will compare how rainfall and NDVI model predicts mortality in other years based on their Z-scored *NDVI* and rainfall.

```{r m6}
df <- merge(ndvi, rain, by = c("sublocation", "year", "season"))
#df$year_season <- apply(ndvi[, c("year", "season")], 1, function(x) paste(x, collapse="_"))

head(df, n=2)
#Rainfall predicts
df$rain_mortality <- predict(m_r, df) 
y <- aggregate(rain_mortality~year, df, mean, na.rm=T)
temp <- y
x11()
plot(rain_mortality~year, data=y, type="l", ylab="Mortality rate (%)", main="Average annual mortality rate",  axes=FALSE, cex.lab=1.2, ylim=c(5,40), col="blue")
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, 40, 5), cex.axis=1.1)
box()

#NDVI predictions
df$ndvi_mortality <- predict(m_n, df) 
y <- aggregate(ndvi_mortality~year, df, mean, na.rm=T)
temp$ndvi_mortality <- y$ndvi_mortality
lines(ndvi_mortality[order(year)]~year[order(year)], data=y, col="red")
legend("topright", c("Rainfall", "NDVI "), lty=1, col=c("blue", "red"), title = "Prediction type", bty = "n")
text(1984,27,"1984")
text(1989,36,"1989")
text(1998, 6,"1998")
text(2000,32,"2000")
text(2009,25,"2009")
text(2011,29,"2011")

```

Now plot rainfall and NDVI average and compare,

```{r m6b}
x11()
temp <- t(ref)
colnames(temp ) <- ref$year
temp <- temp[-1,-1]
barplot(temp, las=2, col=c("blue", "green"), ylab=" ", xlab="Year")
legend("topright",   legend=c("Rainfall (mm)", "NDVI"), fill=c("blue", "green"), bty = "n",cex=1)

```

### Design

Now that we have a relationship between NDVI and mortality, we can design an insurance contract. Let us assume we design an insurance contract that pays when mortality rate falls below the *80th percentile* of the long term series.

```{r m7}
df <- na.omit(df)
r_trig <- quantile(df$rain_mortality, 0.80, na.rm=T)
round(r_trig, 2)
n_trig <- quantile(df$ndvi_mortality, 0.80, na.rm=T)
round(n_trig, 2)

```

The trigger first occurs at the following *zNDVI* and *z-rainfall*.

```{r m8}
tndvi <- max(df$zndvi[df$ndvi_mortality > n_trig])
round(tndvi, 2)
train <- max(df$zrain[df$rain_mortality > r_trig])
round(train, 2)

```

Now let analyse plots of our designed contract.

```{r m9}
x11()
plot(ndvi_mortality[order(zndvi)]~zndvi[order(zndvi)], data=df, xlab="Z-NDVI", ylab="Predicted mortality", pch=16, type="l", lwd=1.2)
abline(h=n_trig,col="green")
abline(v=tndvi,col="red")
text(1,0.3, "False Positive")
text(-2,0.05, "False Negatives")
x11()
plot(rain_mortality[order(zrain)]~zrain[order(zrain)], data=df, xlab="Z-rainfall", ylab="Predicted mortality", pch=16, type="l", lwd=1.2)
abline(h=r_trig,col="green")
abline(v=train,col="red")
```

The payout — in units of monetary value of TLU — can be computed as simply the percentage difference between predicted mortality in TLU and the trigger. Those below the trigger get no payout. We assume that 1 TLU has a monetary equivalence of 1000$.

```{r m10}
df$r_pay <- pmax(0, df$rain_mortality - r_trig)/100 * 1000
df$n_pay <- pmax(0, df$ndvi_mortality - n_trig)/100 * 1000
```

Plots for payouts

```{r m11}
#Rainfall
y <- aggregate(r_pay~year, df, mean, na.rm=T)
y$r_pay <- y$r_pay[order(y$year)]
y$year <- y$year[order(y$year)]
x11()
plot(r_pay~year, data=y, type="l", ylab="Payouts (USD)", main="Average annual payouts using rain data",  axes=FALSE, cex.lab=1.1)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, round(max(y$r_pay,-1)), by=round(max(y$r_pay)/10,-1) ), cex.axis=1.1)
box()

#NDVI
y <- aggregate(n_pay~year, df, mean, na.rm=T)
y$n_pay <- y$n_pay[order(y$year)]
y$year <- y$year[order(y$year)]
x11()
plot(n_pay~year, data=y, type="l", ylab="Payouts (USD)", main="Average annual payouts using NDVI data",  axes=FALSE, cex.lab=1.2)
axis(side=1, at=seq(1982,2019, 2), cex.axis=1.1)
axis(side=2, at=seq(0, round(max(y$n_pay,-1)), by=round(max(y$n_pay)/10,-1) ), cex.axis=1.1)
box()

x11()
par(mfrow=c(2, 1), mar=c(4.5, 4, 0.5, 0.5)) #c(bottom, left, top, right)
boxplot(n_pay~year, data=df,cex=.25, ylab="NDVI payouts ($)")
boxplot(r_pay~year, data=df,cex=.25, ylab="Rainfall payouts ($)")

```

Let's explore how the payout function behaves.

```{r m12}
#Plot NDVI payout function
x11()
par(mfrow=c(1, 2), mar=c(17, 4, 1, 0.5)) #c(bottom, left, top, right)
plot(n_pay[order(n_pay)]~zndvi[order(n_pay)], data=df, ylab="Payouts ($)", xlab="Z-NDVI", cex.lab=1.2, cex.axis=1.2, pch=16, cex=0.7, type="l")
abline(v=tndvi, col="red")
text(1.0, 400,"Trigger", cex=1.2)
#Plot rainfall payout function
plot(r_pay[order(r_pay)]~zrain[order(r_pay)], data=df, ylab="Payouts ($)", xlab="Z-rainfall", cex.lab=1.2, cex.axis=1.2, pch=16, cex=0.7, type="l", lwd=1.2)
abline(v=train, col="red", lwd=1.2)
text(0.1, 140,"Trigger", cex=1.2)

```

### Evaluation

We now need to assess the welfare of index insurance to the pastoralists. To achieve this, we adopt the Minimum Quality Standard (MQS) test by *Carter and Steinmetz (2018)*. The goal of MQS test is to answer the question; "would a household be better off with or without insurance?" If the household would be better off economically buying insurance, then we will say that the insurance contract meets the MQS. Basically, contracts that fail too often and at the wrong times can make farmers worse off, i.e. a farmer looses both insurance premium and livestock. This is why the MQS test is conducted to evaluate the welfare of agriculture insurance contracts.

The *Certainty Equivalence* (CE) is a key concept in evaluating welfare of insurance programs. It expresses the value of an uncertain stream of income $y$ (one year you earn more, another year you earn less) in terms of a stable income. Most people prefer a stable income to an unstable income, even if that unstable income is (somewhat) higher on average. This is because of having a year with a low income can have a great negative effect on someone's welfare. CE  can be computed using a utility function as,

$$
    U(y) = \begin{cases} y^{(1-\rho)} / {(1-\rho)}    & \text{if } \rho \neq 1 \\
    \ln (y) & \text{if } \rho = 1
    \end{cases}
$$

where $rho$ is a household's/individual's level of risk aversion and higher $rho$ indicates higher levels of risk aversion. Here, *risk aversion* refers to the degree to which households are averse to (avoid) losses relative to gains ([Jensen et. al 2019](https://doi.org/10.1016/j.ecolecon.2019.04.014)). For example, a risk averse person is one who prefers lower returns with known risk rather than higher returns with unknown risks. Essentially, a utility function expresses the diminishing returns to increased income. For instance, a pastrolist derives more additional utility if their income increases from $100 to $300 than if goes from $4000 to $4300. How much exactly?, is hard to say. It depends on your preferences and circumstances. Therefore, the utility function is an important concept that measures *preferences* over a set of rewarding outcomes. The economist's utility function is thus a *weighting function* that tells us how much we dread the bad state of the world. Therefore, given a stream of income of a pastrolist CE can be determined as

$$
    ce(y) = \begin{cases} \exp(\sigma)    & \text{if } \rho = 1 \\
     (1 - \rho)^{\frac{1}{1 - \rho}} & \text{otherwise}
    \end{cases}
$$

where $\sigma=E[u(y)]$ i.e. expected utility.

We finally determine the welfare of insurance using the MQS test based on CE with insurance $ce(y_i)$ and without $ce(y)$ as:

$$
    mqs(y) = \begin{cases} \text{if }  ce(y_i) < ce(y)  & \text{Bad } \\
     \text{if }  ce(y_i) = ce(y) & \text{Neutral}\\
     \text{if }  ce(y_i) > ce(y) & \text{Good}
    \end{cases}
$$

The MQS has 3 outcomes based on the insurance contract i.e. negative value, zero, and positive which corresponds to bad, neutral and good respectively. Remember, the MQS test goal is to answer the question; "**would a farmer be better off with insurance or without insurance?**" If the farmer would be better off economically buying insurance, then we will say that the insurance contract meets the **Minimum Quality Standard (MQS)** hence positive value.

Prepare data.

```{r e1}
data <- df[df$year >=2008,]
temp <- na.omit(d[, c("sublocation", "season", "year", "mortality_rate", "stock_beginning")])
data <- merge(data, temp, by=c("sublocation", "season", "year"))
```

Compute pastrolist capital as the percentage of TLU remaining after losses. NB: 1 TLU is $1000.

```{r e2}
data$capital <- (1 - data$mortality_rate) * 1000
```


Let us assess the welfare of index insurance to the pastoralists. So get the CE function.

```{r e3}
ce_income
```

We will compute income with and without insurance. The Actuary Fair Premium (AFP) is given us as average indemnities $I$ i.e. 
$$
 \text{afp} = \sum_{i=1}^{n} \frac{I_i}{n}
$$
where $n$ is the number of payouts.

In this case the market price of the premium includes some mark-up $m$ for to cater for administrative cost, hence the premium $P$ is given as:

$$
 \text{P} = \text{afp} \times (1 + m)
$$

Assume the premium that is paid for IBLI insurance as equivalent to the expected payouts with a mark up of 25%. We will basically compare the IBLI welfare based on the observed mortality losses and predicted mortality based on rainfall and NDVI.

```{r e4}
markup <- 0.25
#NDVi
n_premium <- mean(data$n_pay) * (1 + markup)
n_premium
df1 <- data.frame(y_noins = data$capital)
df1$y_nomarkup <- (data$capital + data$n_pay) - mean(data$n_pay)
df1$y_25markup <- (data$capital + data$n_pay) - n_premium

#Rainfall model premiums
r_premium <- mean(data$r_pay) * (1 + markup)
r_premium
df2 <- data.frame(y_noins = data$capital)
df2$y_nomarkup <- (data$capital + data$r_pay) - mean(data$r_pay)
df2$y_25markup <- (data$capital + data$r_pay) - r_premium

```

Let us compute CE over $0 \leq \rho \leq 10$ and plot CE pay in USD when there is no mark up on insurance (*ce_ins_nomarkup*), when insurance is 25% marked up (*ce_ins*) and a case where no insurance is taken by pastrolist (*ce_base*).

```{r e6}
ce <- function(rhos, ce_base, ce_ins, ce_ins_nomarkup, df1){
  for(i in 1:length(rhos)){
    ce_base[i] <- ce_income(df1$y_noins, rhos[i])
    ce_ins[i]  <- ce_income(df1$y_25markup, rhos[i])
    ce_ins_nomarkup[i] <- ce_income(df1$y_nomarkup, rhos[i])
  }
  return(list(ce_base=ce_base, ce_ins=ce_ins,ce_ins_nomarkup=ce_ins_nomarkup))
}

rhos <- seq(0, 5, .1)
#NDVI
ce_base <- ce_ins <- ce_ins_nomarkup <- rep(NA,length(rhos))
n_ce <- ce(rhos, ce_base, ce_ins, ce_ins_nomarkup, df1)

#RAINFALL
ce_base <- ce_ins <- ce_ins_nomarkup <- rep(NA,length(rhos))
r_ce <- ce(rhos, ce_base, ce_ins, ce_ins_nomarkup, df2)

x11()
plot(rhos, n_ce$ce_base, type="l", ylab= "Certainty Equivalent (USD per TLU)", xlab="CRRA", cex=2)
lines(rhos, n_ce$ce_ins, col="blue")
lines(rhos, n_ce$ce_ins_nomarkup, col="green")
lines(rhos, r_ce$ce_ins, col="red")
lines(rhos, r_ce$ce_ins_nomarkup, col="magenta")
legend("bottomleft", c("No insurance", "NDVI 25% Marked-up", "NDVI No mark up", "Rainfall 25% Marked-up", "Rainfall No mark up"), lty=1, col=c("black", "blue", "green", "red","magenta"), title = "Insurance type", bty = "n")

```

We can observe that insurance with no mark up is always beneficial. On the other hand, insurance that is marked-up starts being beneficial for persons with risk aversion factor above 1.5. Basically, risk neutral pastrolist would not derive any benefit from this insurance contract.

Let do the *MQS test* using the same values of CRRA and make a plot to determine pastrolist welfare with and without insurance.

```{r e7}
test <- function(rhos, df1){
  ce_base <- ce_ins <- mqs <- mqs_no <- rep(NA, length(rhos))
    for(i in 1:length(rhos)){
      ce_base[i] <- ce_income(df1$y_noins, rhos[i])
      ce_ins[i]  <- ce_income(df1$y_25markup, rhos[i])
      ce_ins_nomarkup[i] <- ce_income(df1$y_nomarkup, rhos[i])
      mqs[i]    <- ce_ins[i] - ce_base[i]
      mqs_no[i] <- ce_ins_nomarkup[i] - ce_base[i]
    }
  return(list(mqs=mqs,mqs_no=mqs_no))
}

#NDVI
mqs_n <- test(rhos, df1)
#Rainfal
mqs_r <- test(rhos, df2)

x11()
plot(rhos, mqs_n$mqs, type="l", col="red", ylab= "MQS (insurance benefit per TLU in USD)", xlab="CRRA", cex=2, ylim=c(-10, 60))
lines(rhos, mqs_n$mqs_no, col="green")
lines(rhos, mqs_r$mqs_no, col="magenta")
lines(rhos, mqs_r$mqs, col="blue")
abline(h=0, col="black", cex=0.6)
legend("topleft", c("NDVI 25% Marked-up", "NDVI No mark up", "Rainfall No mark up", "Rainfall 25% Marked-up"), lty=1, col=c("red", "green", "magenta", "blue"), title = "Insurance type", bty = "n")

#text(3, 5, "Good", cex=1.2)
#text(3, 0, "Neutral", cex=1.2)
#text(3, -3, "bad", cex=1.2)
```

### Simulation

On the above experiments we have used observed mortality rates (2008--2014) to estimate capital/income for a pastrolist per TLU. In principle we assumed that each farmer had 1 TLU in the absence of loss. We then proceeded to evaluate insurance contracts using predicted mortality rates. However, there is uncertainty that is associated with our prediction model. Therefore we conduct a simulation of pay based on 95% confidence interval of the model’s predictions.

Create a model that predicts mortality from zNDVI and rainfall with variability bands.

```{r s1}
#NDVI
m_n <- loess.sd(data$zndvi, data$mortality_rate, control=loess.control(surface="direct"))
fitm_n <- cbind(fit=m_n$model$fitted, sd=m_n$sd)
#Rainfall
m_r <- loess.sd(data$zrain, data$mortality_rate, control=loess.control(surface="direct"))
fitm_r <- cbind(fit=m_r$model$fitted, sd=m_r$sd)
```


Let us randomly draw 180 normally distributed samples a 1000 times from the model's distribution we created previously.

```{r s2}
ns <- 1000
#NDVI
n_sample <- apply(fitm_n, 1, function(i) {
  pmin(1, pmax(0, rnorm(ns, i[1], i[2])))
})
n_sample <- t(n_sample)

#Rainfall
r_sample <- apply(fitm_r, 1, function(i) {
  pmin(1, pmax(0, rnorm(ns, i[1], i[2])))
})
r_sample <- t(r_sample)

```


Let us now compute the MQS test using random draws from rainfall and NDVI distribution.

```{r s3}
rhos <- seq(0,5,1)
simulate_mqs <- function(rhos, ns, payouts, premium, samples){
  mqs <- mqs_no <- rep(NA, length(rhos))
  for(j in 1:length(rhos)){
    out_nomarkup <- out_ins <- out_base <- rep(NA, ns)
    for (i in 1:ns) {
      rho <- rhos[j]
      y_noins <- (1 - samples[,i]) * 1000 #Capital or base $1000 per TLU
      insurance_nomarkup <- y_noins + payouts
      insurance <- insurance_nomarkup - premium
      out_base[i] <- ce_income(y_noins, rho)
      out_ins[i]  <- ce_income(insurance, rho)
      out_nomarkup[i]  <- ce_income(insurance_nomarkup, rho)
    }
    mqs[j]     <-  mean(out_ins - out_base)
    mqs_no[j]  <-  mean(out_nomarkup - out_base)
  }
  return(list(mqs=mqs,mqs_no=mqs_no))
}
  
#NDVI
mqs_n <- 0
mqs_n <- simulate_mqs(rhos, ns, data$n_pay, n_premium, n_sample)
#Rainfall
mqs_r <- 0
mqs_r <- simulate_mqs(rhos, ns, data$r_pay, r_premium, r_sample)

x11()
plot(rhos, mqs_n$mqs, type="l", col="red", ylab= "MQS (insurance benefit per TLU in USD)", xlab="CRRA", cex=2, ylim=c(-10, 90))
lines(rhos, mqs_n$mqs_no, col="green")
lines(rhos, mqs_r$mqs_no, col="magenta")
lines(rhos, mqs_r$mqs, col="blue")
abline(h=0, col="black", cex=0.6)
legend("topleft", c("NDVI 25% Marked-up", "NDVI No mark up", "Rainfall No mark up", "Rainfall 25% Marked-up"), lty=1, col=c("red", "green", "magenta", "blue"), title = "Insurance type", bty = "n")

```


### References

Carter, Michael and Tara Steinmetz (2018). “Quality Standards for Agricultural Index Insurance: An Agenda for Action,” in *State of Microfinance*, MunichRE.

Jensen, N., Stoeffler, Q., Fava, F., Vrieling, A., Atzberger, C., Meroni, M., Mude, A. & Carter, M. (2019). [Does the design matter? Comparing satellite-based indices for insuring pastoralists against drought](https://doi.org/10.1016/j.ecolecon.2019.04.014). *Ecological economics*, 162, 59-73.

``

