---
title: "Determination of insurance zones"
author: "Benson Kenduiywo"
date: "04/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data preparation

Load libraries
 
```{r lib}
rm(list = ls(all=TRUE))
library(agrins)
library(reshape2)
library(dplyr)
path <- "D:/UCDavis/QUIIC/github/IBLI/"
source(paste0(path,"scripts/Functions.R"))
```
Load data.

```{r d1}
#NDVI
n <- data_ibli("marsabit_avhrr_ndvi") #NOAA
#Precipitation
p <- data_ibli("marsabit_chirps_precipitation")
names(p)[1:10]
#Adim boundary
subloc <- shapefile(paste0(path,"data/marsabit_subloc_ibli.shp"))
names(subloc)[6] <- "sublocation"
#Mortality
mort <- data_ibli("marsabit_mortality.rds")
#Causes of mortality data
cause <- data_ibli("marsabit_losses")[,c("hhid", "sublocation", "cause","month", "year")]
colnames(cause)[4] <- "season"
s <- cause$season
cause$season <- "SRSD"
cause$season[s >=3 & s <=9] <-"LRLD"
cause <- na.omit(cause)
head(cause, n=2)
```

Reshape the data from wide table to long with year as variable.

```{r d2}
#http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/
p2 <- melt(p, variable.name = "date", value.name = "rainfall",id.vars = "SUBLOCATION")
p2[1,]
n2 <- melt(n, variable.name = "date", value.name = "ndvi", id.vars = "SUBLOCATION")
n2[1,]

```

Format the date variable by removing unnecessary variables.

```{r d3}
#Chirps
date <- gsub("X", "", p2$date)
p2$date <- as.Date(as.character(date), format = "%Y%m%d")
p2[1,]
#NOAA
date <- gsub("X", "", n2$date)
n2$date <- as.Date(as.character(date), format = "%Y%m%d")
n2[1,]

```
Merge causes table with mortality data.

```{r d3a}
mort2  <- merge(mort, cause, by=c("hhid","sublocation", "season", "year"))

```
Compute daily average rainfall and NDVI per season in each year.

```{r d4}
years <- seq(1982, 2019, 1)
LR_rain <- lapply(years, seasonMean, p2, "long")
LR_rain <- do.call(rbind, LR_rain)
names(LR_rain)[2] <- "rainfall"
SR_rain <- lapply(years, seasonMean, p2, "short")
SR_rain <- do.call(rbind, SR_rain)
names(SR_rain)[2] <- "rainfall"
rain <- rbind(LR_rain, SR_rain)
#NDVI
LR_ndvi <- lapply(years, seasonMean, n2, "long")
LR_ndvi <- do.call(rbind, LR_ndvi)
names(LR_ndvi)[2] <- "ndvi"
SR_ndvi <- lapply(years, seasonMean, n2, "short")
SR_ndvi <- do.call(rbind, SR_ndvi)
names(SR_ndvi)[2] <- "ndvi"
ndvi <- rbind(LR_ndvi, SR_ndvi)

```

Compute the zScored daily average rainfall per season per sublocation.

```{r d4a}
rain <- ungroup(mutate(group_by(rain,SUBLOCATION), 
                       zrain=zscore(rainfall)))

ndvi <- ungroup(mutate(group_by(ndvi,SUBLOCATION), 
                       zndvi=zscore(ndvi)))
```

Compute mortality.

```{r d5}
#Consider: Disease, Starvation, Rain, predation
mort2 <- mort2[!(mort2$cause %in% c("Just lost", "Old age","Premature birth")),]
mort2 <- mort2[mort2$type == "TLU", ]
mort2 <- na.omit(mort2)
amort <- aggregate(mort2[,c("stock_beginning", "loss"), drop=FALSE], mort2[, c("season","year","sublocation"), drop=FALSE], mean, na.rm=TRUE)
amort$mortality_rate <- amort$loss/amort$stock_beginning

```

Merge mortality with NDVI and rainfall data.

```{r d6}
colnames(ndvi)[1] <- "sublocation"
d <- merge(ndvi, amort, by=c("year", "season", "sublocation"))
colnames(rain)[1] <- "sublocation"
d <- merge(rain, d, by=c("year", "season", "sublocation"))
d$mortality_rate <- d$mortality_rate * 100
```

## Spatial Variation of mortality

Display the spatial variation on mortality.

```{r sv1}
temp <- aggregate(amort[, "mortality_rate", drop=FALSE], amort[, c("season", "sublocation"), drop=FALSE],mean, na.rm=T)
#LRLD
lr <- merge(subloc, temp[temp$season=="LRLD", c("mortality_rate", "sublocation")], by= "sublocation")
#SRSD
sr <- merge(subloc, temp[temp$season=="SRSD", c("mortality_rate", "sublocation")], by= "sublocation")

cols <- heat.colors(length(lr$sublocation), rev=T)
#Rainfal spatial variation
x11()
par(mfrow=c(1, 2), mar=c(1, 1, 2, 2)) #c(bottom, left, top, right)
spplot(lr, "mortality_rate", col.regions=cols, main="LRLD Average observed mortality per Sub-location: 2008-2015")
spplot(sr, "mortality_rate", col.regions=cols, main="SRSD Average observed mortality per Sub-location: 2008-2015")

```

## Clustering zones

Use kmeans to cluster zones based on NDVI and rainfall per season per year

```{r z1}

#Convert long to Wide
#==================================
n_w <- dcast(ndvi, sublocation ~ year+season, value.var="ndvi")
r_w <- dcast(rain, sublocation ~ year+season, value.var="rainfall")
df <- merge(n_w,r_w, by=c("sublocation"))
subs <- sort(unique(d$sublocation))
aoi <- subloc[subloc$sublocation %in% subs,]
df1 <- df[df$sublocation %in% subs,]
set.seed(99)
#rf <-  randomForest(df[,-1],ntree=200,importance = TRUE)
#Create 10 clusters using 500 iterations using 5 random sets from "Lloyd" method 
r_kmn <- kmeans(df1[,-1], centers = 10, iter.max = 500, nstart = 5, algorithm="Lloyd")
# kmeans returns an object of class "kmeans"
aoi$Zones <- r_kmn$cluster

```
Display the old and new zones

```{r z2}
x11()
plot(subloc, main= "NDVI+Rain")
text(aoi, aoi$Zones, cex=.8)
#Merge to current boundaries
z_10 <- aggregate(aoi, by="Zones")
plot(z_10, add = TRUE, border = "red", lwd = 2)

```

## Effect of zones on mortality predictions

We will determine zones between 4--10 and evaluate the effect of the zones on mortality predictions based on R$^2$, Mean Absolute Prediction Error (MAPE) and RMSE. For instance, a MAPE of (1) < 10 indicates highly accurate prediction, while (2) 10--0 good prediction,
 3) 20 -- reasonable prediction, and (4) > 50 inaccurate.

```{r e1}
zones <- seq(2,15,1) #c(4:10)
r_n <- c()
e_n <- c()
r_r <- c()
e_r <- c()
mape_n <- c()
mape_r <- c()
#Construct a Global mortality model based on NDVI & rainfall
m_r <- loess(mortality_rate ~ zrain , data=d, control=loess.control(surface="direct"))
m_n <- loess(mortality_rate ~ zndvi, data=d, control=loess.control(surface="direct"))

for(z in 1:length(zones)){
  set.seed(99)
  r_kmn <- kmeans(df1[,-1], centers = zones[z], iter.max = 500, nstart = 2, algorithm="Lloyd")
  aoi$Zones <- r_kmn$cluster
  df2 <- 0
  df2 <- merge(d, aoi[,-c(1:5,7:12)], by="sublocation")
  df2 <- ungroup(mutate(group_by(df2, season, year,Zones), 
                       zrain=mean(zrain), zndvi=mean(zndvi)))
  #Rainfall
  p <- predict(m_r, df2)
  r_r[z] <- R_square(df2$mortality_rate, p)
  e_r[z] <- rmse(df2$mortality_rate - p)
  mape_r[z] <- MAPE(df2$mortality_rate, p)
  #NDVI
  p <- predict(m_n, df2)
  r_n[z] <- R_square(df2$mortality_rate, p)
  e_n[z] <- rmse(df2$mortality_rate - p)
  mape_n[z] <- MAPE(df2$mortality_rate, p)
}
quality <- data.frame(No_Zone=zones,ndvi_R2=r_n,rain_R2=r_r, ndvi_RMSE=e_n,rain_RMSE=e_r,ndvi_mape=mape_n,rain_mape=mape_r)
quality

```

Plot the influence of zone size on R$^2$ and RMSE of the mortality model.

```{r e2}
windows()
#par(mfrow=c(1,2),mar=c(15,4.5,1,1))
par(mfrow=c(1, 2), mar=c(15, 4.5, 2, 0.5)) #c(bottom, left, top, right)
plot(zones, r_n, type="l", ylim= c(0,1), col="green", ylab=expression(R^2), xlab= "Sub-location clusters", cex.axis =1.2, cex.lab = 1.2, lwd=2)
lines(zones, r_r, col="blue", lwd=2)
legend("topright", c("NDVI", "Rainfall"), col=c("green","blue"), lty = 1, lwd=2)

plot(zones, mape_n, type="l", ylim= c(0,100), col="green", ylab="MAPE (%)", xlab= "Sub-location clusters", cex.axis =1.2, cex.lab = 1.2,lwd=2)
lines(zones, mape_r, col="blue",lwd=2)
legend("topright", c("NDVI", "Rainfall"), col=c("green","blue"), lty = 1, lwd=2)

```

## Influence of number zones on MQS

We evaluate the influence of zone clusters on insurance welfare using MQS based on a risk aversion coefficient $\rho=1.5$ -- which falls within the common range for risk-averse farming communities.

```{r mqs1}
markup <- 0.25
rhos <- 1.5 #seq(0, 5, .1)
mqs_n <- mqs_r <- vector("list", length(rhos))
for(z in 1:length(zones)){
  set.seed(99)
  r_kmn <- kmeans(df1[,-1], centers = zones[z], iter.max = 500, nstart = 2, algorithm="Lloyd")
  aoi$Zones <- r_kmn$cluster
  df2 <- 0
  df2 <- merge(d, aoi[,-c(1:5,7:12)], by="sublocation")
  df2 <- ungroup(mutate(group_by(df2, season, year,Zones), 
                       zrain=mean(zrain), zndvi=mean(zndvi)))
  #1.0 Mortality model
  #Rainfall
  df2$rain_mortality <- predict(m_r, df2)
  #NDVI
  df2$ndvi_mortality <- predict(m_n, df2)
  #2.0 Contract design
  r_trig <- quantile(df2$rain_mortality, 0.80, na.rm=T)
  n_trig <- quantile(df2$ndvi_mortality, 0.80, na.rm=T)
  #make payouts; 1 TLU=100$
  df2$r_pay <- pmax(0, df2$rain_mortality - r_trig)/100 * 1000
  df2$n_pay <- pmax(0, df2$ndvi_mortality - n_trig)/100 * 1000
  #3.0 Evaluation
  #NDVI
  df2$capital <- (100 - df2$mortality_rate) * 1000
  n_premium <- mean(df2$n_pay) * (1 + markup)
  df_n <- data.frame(y_noins = df2$capital) # no insurance
  df_n$y_nomarkup <- (df2$capital + df2$n_pay) - mean(df2$n_pay) #no markup
  df_n$y_25markup <- (df2$capital + df2$n_pay) - n_premium #marked up
  #Rainfall model premiums
  r_premium <- mean(df2$r_pay) * (1 + markup)
  df_r <- data.frame(y_noins = df2$capital)
  df_r$y_nomarkup <- (df2$capital + df2$r_pay) - mean(df2$r_pay)
  df_r$y_25markup <- (df2$capital + df2$r_pay) - r_premium
  #4.0 MQS test
  mqs_n[[z]] <- test(rhos, df_n)
  mqs_r[[z]] <- test(rhos, df_r)

}

```

We have made a welfare test on the various zones we have created. Let us now make a plot to evaluate if the size of the zone clusters influence insurance welfare in any way.

```{r mqs2}
a <-  matrix(unlist(mqs_n),ncol=2,byrow=TRUE)
b <-  matrix(unlist(mqs_r),ncol=2,byrow=TRUE)
x11()
plot(zones, a[,1], type="l", col="red", ylab= "MQS (insurance benefit per TLU in $)", xlab="Sub-location clusters", cex=2, ylim=c(0, 30), main=expression(paste(rho," = 1.5")))
lines(zones, a[,2], col="green")
lines(zones, b[,2], col="magenta")
lines(zones, b[,1], col="blue")
legend("topright", c("NDVI 25% Marked-up", "NDVI No mark up", "Rainfall No mark up", "Rainfall 25% Marked-up"), lty=1, col=c("red", "green", "magenta", "blue"), title = "Insurance type", bty = "n")

```

Consider a highly risk averse farmer $\rho=5$

```{r mqs3}
markup <- 0.25
rhos <- 5 #seq(0, 5, .1)
mqs_n <- mqs_r <- vector("list", length(rhos))
for(z in 1:length(zones)){
  set.seed(99)
  r_kmn <- kmeans(df1[,-1], centers = zones[z], iter.max = 500, nstart = 2, algorithm="Lloyd")
  aoi$Zones <- r_kmn$cluster
  df2 <- 0
  df2 <- merge(d, aoi[,-c(1:5,7:12)], by="sublocation")
  
  df2 <- ungroup(mutate(group_by(df2, season, year,Zones), 
                       zrain=mean(zrain), zndvi=mean(zndvi)))
  #1.0 Mortality model
  #Rainfall
  df2$rain_mortality <- predict(m_r, df2)
  #NDVI
  df2$ndvi_mortality <- predict(m_n, df2)
  #2.0 Contract design
  r_trig <- quantile(df2$rain_mortality, 0.80, na.rm=T)
  n_trig <- quantile(df2$ndvi_mortality, 0.80, na.rm=T)
  #make payouts; 1 TLU=100$
  df2$r_pay <- pmax(0, df2$rain_mortality - r_trig)/100 * 1000
  df2$n_pay <- pmax(0, df2$ndvi_mortality - n_trig)/100 * 1000
  #3.0 Evaluation
  #NDVI
  df2$capital <- (100 - df2$mortality_rate) * 1000
  n_premium <- mean(df2$n_pay) * (1 + markup)
  df_n <- data.frame(y_noins = df2$capital) # no insurance
  df_n$y_nomarkup <- (df2$capital + df2$n_pay) - mean(df2$n_pay) #no markup
  df_n$y_25markup <- (df2$capital + df2$n_pay) - n_premium #marked up
  #Rainfall model premiums
  r_premium <- mean(df2$r_pay) * (1 + markup)
  df_r <- data.frame(y_noins = df2$capital)
  df_r$y_nomarkup <- (df2$capital + df2$r_pay) - mean(df2$r_pay)
  df_r$y_25markup <- (df2$capital + df2$r_pay) - r_premium
  #4.0 MQS test
  mqs_n[[z]] <- test(rhos, df_n)
  mqs_r[[z]] <- test(rhos, df_r)

}

```
Make a plot to evaluate if the size of the zone clusters influence insurance welfare in any way.

```{r mqs4}
a <-  matrix(unlist(mqs_n),ncol=2,byrow=TRUE)
b <-  matrix(unlist(mqs_r),ncol=2,byrow=TRUE)
x11()
plot(zones, a[,1], type="l", col="red", ylab= "MQS (insurance benefit per TLU in $)", xlab="Sub-location clusters", cex=2, ylim=c(0, 300), main=expression(paste(rho," = 5")))
lines(zones, a[,2], col="green")
lines(zones, b[,2], col="magenta")
lines(zones, b[,1], col="blue")
legend("topright", c("NDVI 25% Marked-up", "NDVI No mark up", "Rainfall No mark up", "Rainfall 25% Marked-up"), lty=1, col=c("red", "green", "magenta", "blue"), title = "Insurance type", bty = "n")

```


## References

Lewis, C. D., 1982. *Industrial and business forecasting methods:
a practical guide to exponential smoothing and curve fitting*.
Butterworth Scientific, London Boston.
