---
title: "IBLI contract design and evaluation based on chirps and NDVI"
author: "Benson Kenduiywo"
date: "22/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## IBLI Programme Design and Evaluation using CHIRPS rainfall data

 Load libraries
 
```{r lib}
rm(list = ls(all=TRUE))
library(agrins)
library(reshape2)
library(dplyr)

```

Load precipitation data.

```{r d1}
p <- data_ibli("marsabit_chirps_precipitation")
names(p)[1:10]

```

Reshape the data from wide table to long with year as variable.

```{r d2}
p2 <- melt(p, variable.name = "date", value.name = "rainfall", id.vars = "SUBLOCATION")
p2[1,]

```

Format the date variable by removing unnecessary variables.

```{r d3}
date <- gsub("X", "", p2$date)
p2$date <- as.Date(as.character(date), format = "%Y%m%d")
p2[1,]

```

Let us make some plots.

```{r plots}
plot(rainfall~date, data=p2, type="l", ylab="rainfall (mm)")
m <- months(p2$date)
i <- m == "September"
lines(p2$date[i], p2$rainfall[i], col="red", cex=1.2)
j <- m == "October"
lines(p2$date[j], p2$rainfall[j], col="blue", cex=1.2)
legend("topleft", lty = 1, col = c("red", "blue"), legend=c('September','October'), title="Month")

#1980-2019 Average monthly rainfall
p2$months <- months(p2$date)
m <- aggregate(rainfall~months, p2, mean, na.rm=T)
m$months <- as.Date(paste0("2020", m$months,"01"), format = "%Y%B%d")

m <- m[order(m$months),]
plot(rainfall~months, data=m, type="l", ylab="rainfall (mm)", main="1980-2019 Average monthly rainfall")

#1980-2019 average annual rainfall
p2$year <- as.numeric(format(p2$date,'%Y')) 
y <- aggregate(rainfall~year, p2, mean, na.rm=T)
plot(rainfall~year, data=y, type="l", ylab="rainfall (mm)", main="1980-2019 average annual rainfall")

```

Create a function to compute means for Long Rain Long Dry (March -- September) and Short Rain Short Dry (October -- February) seasons.

```{r d4}
seasonMean <- function(year, df, season){
  if(season =="long"){
    sdate <- paste0(year, "-03-01")
    edate <- paste0(year, "-09-30")
    season <- "LRLD"
  }else if (season =="short"){
    sdate <- paste0(year-1, "-10-01")
    edate <- paste0(year, "-02-28")
    season <- "SRSD"
  }else{
    print("Define season")
  }
  ydf <- df[df$date >= sdate & df$date <= edate, ]
  ym <- aggregate(ydf[,3], ydf[,1, drop=FALSE], mean, na.rm=T)
  ym$year <- year
  ym$season <- season
  return(ym)
}

```

Compute the season daily average rainfall per season in each year.

```{r d5}
years <- seq(1982, 2018, 1)
LR_rain <- lapply(years, seasonMean, p2, "long")
LR_rain <- do.call(rbind, LR_rain)
names(LR_rain)[2] <- "rainfall"
SR_rain <- lapply(years, seasonMean, p2, "short")
SR_rain <- do.call(rbind, SR_rain)
names(SR_rain)[2] <- "rainfall"
rain <- rbind(LR_rain, SR_rain)

#1982-2018 average annual rainfall per season
y <- aggregate(rainfall~year+season, rain, mean, na.rm=T)
plot(rainfall~year, data=y, type="n", ylab="rainfall (mm)", main="1982-2018 average annual rainfall per seaon")
i <- y[y$season=="LRLD", ]
lines(rainfall~year, i, col="red", cex=1.2)
i <- y[y$season=="SRSD", ]
lines(rainfall~year, i, col="blue", cex=1.2)
legend("topleft", lty = 1, col = c("red", "blue"), legend=c('LRLD','SRSD'), title="Season")

```

Compute zScored average daily rainfall per season per Sub-location. First get the zScore function.

```{r d6}

zscore <- function(y){
  (y - mean(y, na.rm=TRUE) ) / (sd(y, na.rm=TRUE))
}

```

The compute the zScored daily average rainfall per season per sublocation.

```{r d7}
rain <- ungroup(mutate(group_by(rain,SUBLOCATION), 
                       zrain=zscore(rainfall)))

```

Now we need house hold mortality data in order to build a model. Get mortality data.

```{r d8}
mort <- data_ibli("marsabit_mortality.rds")

```

Load data with causes of mortality.

```{r d9}
cause <- data_ibli("marsabit_losses")[,c("hhid", "sublocation", "cause","month", "year")]
colnames(cause)[4] <- "season"
s <- cause$season
cause$season <- "SRSD"
cause$season[s >=3 & s <=9] <-"LRLD"
cause <- na.omit(cause)
head(cause, n=2)

```


Merge causes table with mortality data.

```{r d10}
mort2  <- merge(mort, cause, by=c("hhid","sublocation", "season", "year"))

```

Make some plots of cause of deaths.

```{r d11}
temp <- mort2[mort2$type=="TLU", ]
c <- tapply(temp$loss, temp$cause, mean, na.rm=TRUE)
barplot(c)
c <- aggregate(loss~cause+year, temp, mean, na.rm=T)
c <- reshape(c, direction="wide", idvar="cause", timevar="year")
colnames(c) <- gsub("loss.", "", colnames(c))
c$cause <- gsub("/.*","", c$cause)
c$cause <- gsub(" .*","", c$cause)
c$cause[c$cause=="Just"]= "Lost"
c <- as.matrix(c)
rownames(c) <- c[,1]
cola <- c("red", "green", "blue", "grey40", "yellow", "black", "orange", "purple", "brown", "skyblue")
par(mfrow=c(1, 1), mar=c(5, 5, 4, 5))
barplot(c[,-1], legend =  rownames(c), las=2, args.leg=list(x = "topright", bty = "n", inset=c(-0.2, 0),cex=1), col=cola, ylab="Livestock loss per TLU")

```

## Spatial Variation
 Let us explore if there is any spatial variation of average rainfall since 1980.

Load Marsabit shapefile

```{r sp1}
subloc <- data_ibli("marsabit")
```

Check names consistency before merging.

```{r sp2}
a = sort(unique(subloc$SUB_LOCATION))
b = sort(unique(rain$SUBLOCATION))
a[!a %in% b]

```

Now format sublocation names in shapefile to be consistent with those in rain.

```{r sp3}
subloc$SUB_LOCATION <- gsub("\\s*\\([^\\)]+\\)","",subloc$SUB_LOCATION)
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MARSABIT TOWNSHIP"] <- "TOWNSHIP"
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MARSABIT TOWNSHIP"] <- "TOWNSHIP"
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="MT. KULAL" ] <- "MT KULAL" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="LONYORIPICHAU" ] <- "LONYORI PICHAU" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="EL-HADI" ] <- "EL-HARDI" 
subloc$SUB_LOCATION[subloc$SUB_LOCATION=="KURUNGU" ] <- "KARUNGU"

```

Now merge rainfall data with sublocation boundary information and plot LRLD and SRSD spatial variation maps.

```{r sp4}
r <- aggregate(rain[, c("zrain","rainfall"), drop=FALSE], rain[,c("SUBLOCATION", "season"), drop=FALSE],mean, na.rm=T)
lr <- merge(subloc, r[r$season=="LRLD",], by.x="SUB_LOCATION", by.y="SUBLOCATION")
sr <- merge(subloc, r[r$season=="SRSD",], by.x="SUB_LOCATION", by.y="SUBLOCATION")

cols <- terrain.colors(length(lr$SUB_LOCATION),rev = T)
x11()
spplot(lr, "rainfall", col.regions=cols, main="LRLD Average rainfal per Sub-location: 1982-2019")
x11()
spplot(sr, "rainfall", col.regions=cols, main="SRSD Average rainfall per Sub-location: 1982-2019")

```

## Modelling Mortality

### Introduction

To design and evaluate an index based insurance contract, we need to understand the relationship between the index and losses incurred. Here we develop such a relationship using the z-scored rainfall and NDVI to predict livestock mortality, using survey data from Marsabit.
